### 2009-08-28 paper/2009-09/method.tex
Breaking paper into manageable chunks

% Hull description
% This module takes a list of two-dimensional points and reorders them by
% doing multiple convex hull computations. Convex hull is the boundary of
% the minimal convex set containing a given non-empty finite set of points
% in the plane. In other words, all points not in the convex hull are
% enclosed in the convex hull polygon. At each step the convex hull points
% are taken out of the input list and are put into the output list. The
% computation terminates when there are no more points left in the input
% list.

\documentstyle[epsf,11pt]{article}

\begin{document}

\begin{Large}\begin{center}
The Cowichan Problems: \\
Assessing the Usability of Parallel Programming Systems
\end{center}\end{Large}

\begin{center}
\begin{tabular}{cccc}
Gregory V.~Wilson\footnote{\verb`http://www.third-bit.com`}
&
Andrew Borzenko
&
Cameron Gorrie
&
Zachary Kincaid
\end{tabular}
\end{center}

\begin{abstract}
Most programmers believe that parallel programming is harder than sequential programming.
Many also believe that today's parallel programming systems make it even harder.
We have developed a suite of chained toy problems that can be used to assess
the usability of different scientific programming languages, parallel or otherwise.
This paper motivates and describes the suite,
and discusses what we have learned from porting it to four parallel C++ systems
(MPI, OpenMP, Intel Thread Building Blocks, and a Linux tuple space system),
and from translating it into sequential and parallel Haskell.
\end{abstract}

\section{Introduction\label{s:intro}}

Twenty years ago,
Bal {\em{et~al.}\/} listed more than 300 parallel programming systems \cite{b:par-lang-survey}.
A similar survey today could well uncover twice as many,
but a poll of application programmers would probably find that
only a handful are in general use.

The reasons why so many of these PPSs are only used by their developers (if at all) are murkier.
Part of the blame undoubtedly lies with the diversity of hardware,
and the attendant lack of portability,
alluded to above.
A second reason is that an academic researcher's principal interest is developing new languages,
not supporting existing ones.
By the time a system is 90\% complete,
its author is already thinking about its successor.
Few application programmers are willing to invest effort porting codes
to systems which are not properly supported,
and liable to change or evaporate without notice.
Third,
many research groups have chosen to explore the potential of parallelism
without consideration for compatibility with older languages.
Application programmers must therefore face the prospect of completely re-writing their programs,
rather than simply porting them.
As the next section argues,
this can obviate any performance that might be gained from parallelization.
Finally,
many PPSs simply cannot support the software engineering load of large applications.
Many provide only a single global namespace,
do not include I/O capabilities,
require users to marshal and unmarshal data explicitly,
or do not permit hardware to be timeshared.

The net result is that many PPSs are never used for anything more complicated than
generating the Mandelbrot Set,
simulating the Game of Life,
multiplying dense matrices,
or finding all solutions to the $N$-queens problem for some small $N$.
Lacking the feedback that comes from more demanding use,
few systems are improvements over their predecessors
in the way that successive versions of operating systems have improved on one another,
or the languages in the Algol$\rightarrow$Modula-$n$ family
have become both more powerful and more secure.
Most applications programmers therefore choose to play it safe
and stick to parallel programming's lowest common denominators:
data-parallel Fortran,
and message-passing libraries.

The historical reasons for tolerating this situation are becoming less valid with each passing year.
When only a few parallel computers existed,
and these were of widely different types,
software standardization simply didn't make sense.
Similarly,
when supercomputers were very expensive,
the cost of software development seemed relatively smaller,
and the requirement for maximum performance relatively greater,
than in sequential computing.
As \cite{b:social-limits-speed} points out,
however, these arguments are losing force.
Now,
as in sequential computing,
the first question is not ``How do I write programs for this machine?'',
but ``Will this machine run my programs?''
As microprocessor-based MIMD hardware comes to dominate massively-parallel computing,
we see increasing emphasis on standards such as High-Performance Fortran (HPF) \cite{b:hpf-overview},
and the Message-Passing Interface (MPI) \cite{b:mpi-overview}.

But herein lies a danger.
History shows that it is extremely difficult
to displace a programming language or system once it becomes established.
Cobol and MS-DOS are the most frequently cited examples,
but many systems which should have been superceded long ago have managed
not only to cling to life,
but to grow in influence.
(The fact that the mail signature
``The last good thing written in C was Schubert's Ninth Symphony''
can elicit smiles simply proves that even the truth can be funny.)
Established systems evolve slowly, if at all.
It would therefore seem that the best time to guarantee ourselves
a good parallel programming paradigm is right now,
before a large number of large applications are created and take on lives of their own.

We therefore propose to use a set of simple programming problems to compare
the usability of parallel programming systems.
Competent programmers, fluent in a specific system,
will implement solutions to these problems and report their experiences
in terms of development time, code size and clarity, and runtime efficiency.
These problems are on the scale of the toy applications described several paragraphs ago
both to keep implementation time low
and because we feel that any PPS produced today ought to be able to support such simple programs.
As the second author put it,
``We're all sick of seeing these,
but maybe if we all catch the disease at the same time we'll finally build up some immunity to it.''
One significant innovation in this work is that
we also require implementations of our suite to chain problems together,
so that the output of one problem is the input to another.
This will allow us to assess how well each PPS supports modularization and code re-use
(or, more realistically, how much extra effort programmers must invest
to achieve these goals when using a particular PPS).

Section~\ref{s:corollary} motivates this work by showing
how poor usability can limit the realizable parallelism in a program.
Section~\ref{s:previous} surveys previous work on the comparison of parallel programming systems,
while Section~\ref{s:method} outlines our choice of methods, and the constraints on our work.
Section~\ref{s:toys} then presents the suite of problems we intend to use.
In recognition of the original work done by Feo and others on
the Salishan Problems~\cite{b:salishan}, this suite is called
the Cowichan Problems\footnote{\noindent Pronounced {\em{Cow}\/}-i-chun.
	Like Salishan, the word is a Northwest Coast Indian place name.}.

\subsection*{History and Acknowledgments}

The first author originally developed this suite in the mid-1990s,
inspired both by \cite{b:salishan} and by
discussions with R.\ Bruce Irvin (then at the University of Wisconsin),
Henri Bal (Vrije Universiteit, Amsterdam),
and others.
Early experiences were described in \cite{b:cowichan-ifip} and \cite{b:cowichan-orca}.

\section{Quantifying the Importance of Usability\label{s:corollary}}

According to Amdahl's Law,
the time $t$ required to execute a program has two components:
$t_{seq}$, representing sequential operations that cannot be parallelized,
and $t_{par}$, representing those operations which can.
If we let $\sigma$ be the ``serial fraction'' of the program,
so that $t_{seq} = {\sigma}t$ and $t_{par} = (1-{\sigma})t$,
then the speedup which can be achieved using $P$ processors is:
\begin{eqnarray*}
s(P)	& =	& \frac{t(1)}{t(P)}	\\
	& =	& \frac{t}{{{\sigma}t} + {{(1-{\sigma})t}}/{P}}
\end{eqnarray*}
As $P{\rightarrow}{\infty}$,
the potential speedup is bounded by $1/{\sigma}$.

What is not usually acknowledged is
the influence of programming system usability on
the potential speedup of sequential programs.
As stated above, Amdahl's Law applies only to a single run of a program.
In the real world, programs are first developed, and then run many times.
If we let $T$ be the total running time of a program over its lifetime,
$D_{seq}$ be the time required to develop the sequential components of a program,
and $D_{par}$ be the time required to develop the parallel components
(or parallelize the sequential components),
then the lifetime speedup we can achieve is given by:
\begin{eqnarray*}
S(P)	& =	& \frac{D_{seq} + T(1)}{D_{seq} + D_{par} + T(P)}	\\
	& =	& \frac{D_{seq} + T}{D_{seq} + D_{par} + {{\sigma}T} + {{(1-{\sigma})T}}/{P}}
\end{eqnarray*}
Assume that we are parallelizing a ``dusty deck'',
i.e.\ that the time $D_{seq}$ has already been invested.
If we let $\phi = D_{par}/T$ be the ratio of parallelization time to total program runtime,
then achievable speedup is limited by:
\begin{eqnarray*}
S_{legacy}(P)	& =	& \frac{T}{D_{par} + {\sigma}T}		\\
		& =	& \frac{1}{{\phi} + {\sigma}}
\end{eqnarray*}
The time required to parallelize a program can therefore be seen
as increasing that program's effective serial fraction.
Unless parallelization time can be substantially reduced,
massively-parallel computing is likely to remain attractive only in those cases in which:
\begin{itemize}
\item	the application is trivially parallel (e.g.\ task-farming a rendering calculation);
\item	the expected total program runtime is very large
	(i.e.\ the program is a package used extensively in a field
	such as computational chemistry or vehicle crash simulation);
	or
\item	cost is not an obstacle (i.e.\ the customer is the NSA).
\end{itemize}

\section{Previous Work\label{s:previous}}

In a 1981 survey article \cite{b:psych-prog-survey}, B.~A.\ Sheil observed that:
\begin{quote}
As practiced by computer science,
the study of programming is an unholy mixture of mathematics,
literary criticism,
and folklore.
\end{quote}
Little has changed since.
Computer scientists pay lip service to the importance of usability,
but are strangely reluctant to try to quantify it,
particularly with respect to the tools they use themselves.
Religious debates about the ``right'' first-year teaching language
or the ``best'' GUI toolkit are common;
systematic use of the experimental method is not.

Most of the experiments which have been done have concentrated on ``programming in the small'' issues,
such as readability and code-pattern recognition
\cite{b:reverse-parsing-2,b:cloze-program-1,b:prog-comprehension-1}.
The little work that has been done on ``programming in the large'' has been anecdotal,
such as \cite{b:compare-ada-c-pascal}.
This book presented various authors' impressions of Pascal, C, and Ada
(before any complete Ada implementations existed).
Its editors formulated a list of questions to use in comparing languages,
but did not use it themselves.
A modified version of this questionnaire,
tailored for parallel programming languages,
is given in Appendix~\ref{s:questions},
and will be used as a guideline in our work.

Two comparative efforts in parallel programming are \cite{b:babb-cases} and \cite{b:salishan}.
The first presented implementations of a simple numerical quadrature program
in more than a dozen different parallel languages
in use on mid-1980s hardware.
The second presented implementations of
the Salishan Problems---Hamming number generation,
isomer enumeration,
skyline matrix reduction,
and a simple discrete event simulator---in
C* \cite{b:dataparallel-c}, Occam~\cite{b:occam}, Ada~\cite{b:ada},
and a variety of dataflow and functional languages.
Both of these books convey the flavor of the languages they describe,
but neither made any effort to compare languages or problem implementations directly.

Inspired by the Salishan Problems,
the first author developed a suite of seven problems \cite{b:cowichan-1}
in order to test the limits of parallel programming systems.
Individual problems were badly load-balanced,
manipulated irregular pointer-based structures,
and so on.
Our experience implementing them was extremely valuable,
as it uncovered many bugs in the programming system used,
and revealed several shortcomings of the system's design.
However,
we also discovered that implementing each problem takes six to eight weeks,
which we feel is prohibitive at this exploratory stage.

@@

\begin{figure}
\epsfysize=0.5\textheight
\begin{center}\mbox{\epsffile{fig/taxonomy.eps}}\end{center}
\caption{A Taxonomy of Parallel Programming Systems\label{f:taxonomy}}
\end{figure}

@@

\begin{figure}
\epsfxsize=12cm
\begin{center}\mbox{\epsffile{fig/chain.eps}}\end{center}
\caption{Chaining\label{f:chaining}}
\end{figure}

@@

\begin{figure}
\epsfxsize=12cm
\begin{center}\mbox{\epsffile{fig/perf-time.eps}}\end{center}
\caption{Performance vs.\ Time\label{f:perf}}
\end{figure}

@@

\begin{figure}
\epsfxsize=6cm
\begin{center}\mbox{\epsffile{fig/perf-complex.eps}}\end{center}
\caption{Comparative Code Complexity\label{f:complex}}
\end{figure}

@@

\section{The Cowichan Problems\label{s:toys}}

This section describes the problems making up the full problem suite.
As discussed above,
the implementation of each is to be used both as a stand-alone program
and as a module in the chained problem sequence described in Section~\ref{s:issues-chain}.
Table~\ref{t:toy-codes} lists the one-character codes sometimes used to identify toys.

\begin{table}
\begin{center}
\begin{tabular}{cl|cl}
Code	& Toy				& Code	& Toy				\\
\hline
E	& elastic net			& O	& outer product			\\
G	& Gaussian elimination		& P	& matrix-vector product		\\
H	& halving shuffle		& R	& random matrix generation	\\
I	& invasion percolation		& S	& successive over-relaxation	\\
L	& Game of Life			& T	& image thresholding		\\
M	& Mandelbrot Set		& V	& vector difference		\\
N	& point normalization		& W	& point value winnowing
\end{tabular}
\caption{Single-Character Codes for Toys\label{t:toy-codes}}
\end{center}
\end{table}

\subsection{{\tt{mandel}}:
	Mandelbrot Set Generation
	\label{s:toys-mandel}}

\begin{figure}
\begin{center}\framebox{\epsffile{fig/mandel.ps}}\end{center}
\caption{The Mandelbrot Set\label{f:mandel}}
\end{figure}

This module generates the Mandelbrot Set
for a specified region of the complex plane
(Figure~\ref{f:mandel}).
The inputs to this module are:
\begin{description}
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix.
\item[{\tt{x0, y0}}:]
	the real coordinates of the lower-left corner
	of the region to be generated.
\item[{\tt{dx, dy}}:]
	the extent of the region to be generated.
\end{description}
Its output is:
\begin{description}
\item[{\tt{matrix}}:]
	an integer matrix containing the iteration count at each point in the region.
\end{description}

Given initial coordinates $(x_0, y_0)$,
the Mandelbrot Set is generated by iterating the equation
\begin{eqnarray*}
x^{\prime}	& =	& x^2 - y^2 + y_0	\\
y^{\prime}	& =	& 2{x}{y} + x_0
\end{eqnarray*}
until either an iteration limit is reached,
or the values diverge.
The iteration limit used in this module is 150 steps;
divergence occurs when $x^2 + y^2$ becomes 2.0 or greater.
The integer value of each element of the matrix is
the number of iterations done.
Note that,
as in all problems,
the output is required to be
independent of the number of processors used.
This requires programmers to be careful about
floating-point roundoff effects.

\subsection{{\tt{randmat}}:
	Random Number Generation
	\label{s:toys-rng}}

This module fills a matrix with pseudo-random integers.
The inputs to this module are:
\begin{description}
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix.
\item[{\tt{s}}:]
	the random number generation seed.
\end{description}
Its output is:
\begin{description}
\item[{\tt{matrix}}:]
	an integer matrix filled with random values.
\end{description}
Note that,
as in all problems,
the output is required to be independent of the number of processors used.
Generating new seed values from the given seed,
and running one copy of the random number generator
on each processor,
is therefore unlikely to qualify as a solution.

\subsection{{\tt{half}}:
	Two-Dimensional Shuffle
	\label{s:toys-half}}

This module divides the values in a rectangular two-dimensional integer matrix into two halves along one axis,
shuffles them,
and then repeats this operation along the other axis.
Values in odd-numbered locations are collected at the low end of each row or column,
while values in even-numbered locations are moved to the high end.
An example transformation is:

\begin{eqnarray*}
\begin{array}{cccc}
a & b & c & d \\
e & f & g & h \\
i & j & k & l
\end{array}
& = &
\begin{array}{cccc}
a & c & b & d \\
i & k & j & l \\
e & g & f & h
\end{array}
\end{eqnarray*}

Note that how an array element is moved depends only on whether its location index is odd or even,
not on whether its value is odd or even.

The inputs to this module are:
\begin{description}
\item[{\tt{matrix}}:]
	an integer matrix.
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix.
\end{description}
Its output is:
\begin{description}
\item[{\tt{matrix}}:]
	an integer matrix containing shuffled values.
\end{description}

\subsection{{\tt{invperc}}:
	Invasion Percolation
	\label{s:toys-invperc}}

Invasion percolation models the displacement of one fluid
(such as oil)
by another
(such as water)
in fractured rock \cite{b:percolation-theory,b:invasion-percolation}.
In two dimensions,
this can be simulated by generating
an $N{\times}N$ grid of random numbers in the range $[1{\ldots}R]$,
and then marking the center cell of the grid as filled.
In each iteration,
one examines the four orthogonal neighbors of all filled cells,
chooses the one with the lowest value
(i.e.\ the one with the least resistance to filling),
and fills it in.
Figure~\ref{f:invperc1} shows the first few steps in this process,
while Figure~\ref{f:invperc2} shows the evolution of the final fractal shape.

\begin{figure}
\begin{footnotesize}
\begin{math}
\begin{array}{ccc}
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & 38 & 39 & 92 & 38 \\
\hline
44 & 29 & \circ & 29 & 77 \\
\hline
61 & 26 & 90 & 35 & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
&
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & 38 & 39 & 92 & 38 \\
\hline
44 & \star & \circ & 29 & 77 \\
\hline
61 & 26 & 90 & 35 & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
&
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & 38 & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & 29 & 77 \\
\hline
61 & \star & 90 & 35 & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
\\
 & & \\
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & 38 & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & \star & 77 \\
\hline
61 & \circ & 90 & 35 & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
&
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & 38 & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & \circ & 77 \\
\hline
61 & \circ & 90 & \star & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
&
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & 38 & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & \circ & 77 \\
\hline
61 & \circ & 90 & \circ & \star \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
\\
 & & \\
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
10 & \star & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & \circ & 77 \\
\hline
61 & \circ & 90 & 35 & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
&
\begin{array}{|c|c|c|c|c|}
\hline
26 & 12 & 72 & 45 & 38 \\
\hline
\star & \circ & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & \circ & 77 \\
\hline
61 & \circ & 90 & \circ & 11 \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
&
\begin{array}{|c|c|c|c|c|}
\hline
26 & \star & 72 & 45 & 38 \\
\hline
\circ & \circ & 39 & 92 & 38 \\
\hline
44 & \circ & \circ & \circ & 77 \\
\hline
61 & \circ & 90 & \circ & \circ \\
\hline
83 & 84 & 18 & 56 & 52 \\
\hline
\end{array}
\end{array}
\end{math}
\end{footnotesize}
\caption{Invasion Percolation\label{f:invperc1}}
\end{figure}

\begin{figure}
\begin{center}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/invperc1.ps}}}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/invperc2.ps}}}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/invperc3.ps}}}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/invperc4.ps}}}
\end{center}
\caption{Fractal Generated by Invasion Percolation\label{f:invperc2}}
\end{figure}

The simulation continues until
some fixed percentage of cells have been filled,
or until some other condition
(such as the presence of trapped regions)
is achieved.
The fractal structure of the filled and unfilled regions
is then examined to determine how much oil could be recovered.
The na\"{\i}ve way to implement this is to repeatedly scan the array;
a more sophisticated, and much faster, sequential technique
is to maintain a priority queue of unfilled cells
which are neighbors of filled cells.
This latter technique is similar to the list-based methods used in some cellular automaton programs,
and is very difficult to parallelize effectively.

The inputs to this module are:
\begin{description}
\item[{\tt{matrix}}:]
	an integer matrix.
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix and mask.
\item[{\tt{nfill}}:]
	the number of points to fill.
\end{description}
Its output is:
\begin{description}
\item[{\tt{mask}}:]
	a Boolean matrix filled with {\tt{true}} (showing a filled cell) or {\tt{false}} (showing an unfilled cell).
\end{description}
Filling begins at the central cell of the matrix
(rounding down for even-sized axes).

\subsection{{\tt{thresh}}:
	Histogram Thresholding
	\label{s:toys-thresh}}

This module performs histogram thresholding on an image.
Given an integer image $I$ and a target percentage $p$,
it constructs a binary image $B$
such that $B_{i,j}$ is set
if no more than $p$ percent of the pixels in $I$ are brighter than $I_{i,j}$.
The general idea is that an image's histogram should have 2 peaks,
one centered around the average foreground intensity,
and one centered around the average background intensity.
This program attempts to set
a threshold between the two peaks in the histogram
and select the pixels above the threshold.
This module's inputs are:
\begin{description}
\item[{\tt{matrix}}:]
	the integer matrix to be thresholded.
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix and mask.
\item[{\tt{percent}}:]
	the minimum percentage of cells to retain.
\end{description}
Its output is:
\begin{description}
\item[{\tt{mask}}:]
	a Boolean matrix filled with {\tt{true}} (showing a cell that is kept) or {\tt{false}} (showing a cell that is discarded).
\end{description}

\subsection{{\tt{life}}:
	Game of Life
	\label{s:toys-life}}

\begin{figure}
\begin{center}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/life1.ps}}}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/life2.ps}}}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/life3.ps}}}
{\epsfxsize=0.2\textwidth\framebox{\epsffile{fig/life4.ps}}}
\end{center}
\caption{Snapshots from the Game of Life\label{f:life}}
\end{figure}

This module simulates the evolution of Conway's Game of Life,
a two-dimensional cellular automaton (Figure~\ref{f:life}).
Its inputs are:
\begin{description}
\item[{\tt{matrix}}:]
	a Boolean matrix representing the Life world.
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix.
\item[{\tt{numgen}}:]
	the number of generations to simulate.
\end{description}
Its output is:
\begin{description}
\item[{\tt{matrix}}:]
	a Boolean matrix representing the world after simulation.
\end{description}
At each time step,
this module must count the number of live ({\tt{true}}) neighbors
of each cell,
using both orthogonal and diagonal connectivity
and circular boundary conditions.
The update rule is simple:
if a cell has 3 live neighbors,
or has 2 live neighbors and is already alive,
it is alive in the next generation.
In any other situation,
the cell becomes, or stays, dead.

\subsection{{\tt{winnow}}:
	Weighted Point Selection
	\label{s:toys-winnow}}

This module converts a matrix of integer values
to a vector of points,
represented as $x$ and $y$ coordinates.
Its inputs are:
\begin{description}
\item[{\tt{matrix}}:]
	an integer matrix, whose values are used as masses.
\item[{\tt{mask}}:]
	a Boolean matrix showing which points are eligible for consideration.
\item[{\tt{nrows, ncols}}:]
	the number of rows and columns in the matrix.
\item[{\tt{nelts}}:]
	the number of points to select.
\end{description}
Its output is:
\begin{description}
\item[{\tt{points}}:]
	a vector of $(x,y)$ points.
\end{description}
Each location where {\tt{mask}} is {\tt{true}}
becomes a candidate point,
with a weight equal to the integer value in {\tt{matrix}} at that location
and $x$ and $y$ coordinates
equal to its row and column indices.
These candidate points are then sorted into increasing order by weight,
and {\tt{nelts}} evenly-spaced points selected to create the result vector.

\subsection{{\tt{norm}}:
	Point Location Normalization
	\label{s:toys-norm}}

This module normalizes point coordinates so that all points lie within the unit square $[0{\ldots}1]{\times}[0{\ldots}1]$.
Its inputs are:
\begin{description}
\item[{\tt{points}}:]
	a vector of point locations.
\item[{\tt{nelts}}:]
	the number of elements in the point vector.
\end{description}
Its output is:
\begin{description}
\item[{\tt{points}}:]
	a vector of normalized point locations.
\end{description}
If $x_{min}$ and $x_{max}$ are the minimum and maximum $x$ coordinate values in the input vector,
then the normalization equation is:
\[	x_{i}^{\prime} = \frac{x_i - x_{min}}{x_{max} - x_{min}}	\]
$y$ coordinates are normalized in the same fashion.

\subsection{{\tt{elastic}}:
	Elastic Net Simulation
	\label{s:toys-elastic}}

This module uses the elastic net algorithm \cite{b:elastic-net} to find an approximate solution to the travelling salesentity problem (TSP).
A closed, circular loop, modelled by a large number of doubly-connected points, is placed in the plane.
This path is slowly deformed until it connects the points representing the cities to be visited (Figure~\ref{f:elastic}).
The rate at which the loop is deformed determines the quality of the final path.

\begin{figure}
\begin{center}
\framebox{\epsffile{fig/elastic1.ps}}
\framebox{\epsffile{fig/elastic2.ps}}
\framebox{\epsffile{fig/elastic3.ps}}
\end{center}
\caption{Relaxation of an Elastic Net\label{f:elastic}}
\end{figure}

Let the (fixed) location of city $i$ be $c_i$, and the (changeable) location of point $j$ on the elastic net be $p_j$.
The change in the location of point $j$ during each iteration of the algorithm is given by:
\begin{equation}
{\delta}{p_j} = {\alpha}\sum_{i}{w_{ij}(c_i - p_j)} ~+~ K{\beta}(p_{j+1} - 2p_j + p_{j-1})
\end{equation}
$\alpha$ and $\beta$ are time-invariant factors relating the strengths of the city-to-point and point-to-point forces,
and $K$ is a time-dependent factor which controls the rate of annealing.
The strength of the force $w_{ij}$ between city $c_i$ and point $p_j$ is given by:
\begin{equation}
w_{ij} = {\phi}({\mid}c_i - p_j{\mid}, K) / \sum_{\ell}\phi({\mid}c_i - p_{\ell}{\mid}, K)
\end{equation}
where $\phi(d, K)$ is the Gaussian function $e^{-{d^2}/{K^2}}$.
Note the normalization of each $w_{ij}$; in order for this algorithm to work properly, the total force being exerted by each city must be the same.
This is accomplished by summing the forces exerted by each city, and then normalizing the city's forces by that total.

The inputs to this module are:
\begin{description}
\item[{\tt{cities}}:]
	a vector of points representing the locations of the cities to be toured.
	The $(x,y)$ coordinates of these cities must be scaled to lie in the unit square $[0{\ldots}1]{\times}[0{\ldots}1]$.
\item[{\tt{n\_cities}}:]
	the number of cities in {\tt{cities}}.
\item[{\tt{n\_net}}:]
	the number of elastic net points to use.
\item[{\tt{n\_iters}}:]
	the total number of iterations to perform.
\item[{\tt{n\_relax}}:]
	the number of iterations to perform with each discrete value of $K$.
	After each {\tt{n\_relax}} iterations, $K$ is reduced by a (constant) pre-defined scaling factor.
\end{description}
Its output is:
\begin{description}
\item[{\tt{net}}:]
	a vector of {\tt{n\_net}} points representing the tour found.
\end{description}

\subsection{{\tt{outer}}:
	Outer Product
	\label{s:toys-outer}}

This module turns a vector containing point positions
into a dense, symmetric, diagonally dominant matrix
by calculating the distances between each pair of points.
It also constructs a real vector whose values are
the distance of each point from the origin.
Inputs are:
\begin{description}
\item[{\tt{points}}:]
	a vector of $(x,y)$ points, where $x$ and $y$ are the point's position.
\item[{\tt{nelts}}:]
	the number of points in the vector, and the size of the matrix along each axis.
\end{description}
Its outputs are:
\begin{description}
\item[{\tt{matrix}}:]
	a real matrix, whose values are filled with inter-point distances.
\item[{\tt{vector}}:]
	a real vector, whose values are filled with origin-to-point distances.
\end{description}
Each matrix element $M_{i,j}$ such that $i \neq j$
is given the value $d_{i, j}$,
the Euclidean distance between point $i$ and point $j$.
The diagonal values $M_{i, i}$ are then set to
{\tt{nelts}} times the maximum off-diagonal value
to ensure that the matrix is diagonally dominant.
The value of the vector element $v_i$ is set to
the distance of point $i$ from the origin,
which is given by $\sqrt{{x_i}^2 + {y_i}^2}$.

\subsection{{\tt{gauss}}:
	Gaussian Elimination
	\label{s:toys-gauss}}

This module solves a matrix equation $AX=V$ for
a dense, symmetric, diagonally dominant matrix $A$
and an arbitrary vector non-zero $V$
using explicit reduction.
(Matrices are required to be symmetric and diagonally dominant in order to guarantee that
there is a well-formed solution to the equation.)
Inputs are:
\begin{description}
\item[{\tt{matrix}}:]
	the real matrix $A$.
\item[{\tt{target}}:]
	the real vector $V$.
\item[{\tt{nelts}}:]
	the number of values in each vector, and the size of the matrix along each axis.
\end{description}
Its output is:
\begin{description}
\item[{\tt{solution}}:]
	a real vector containing the solution $X$.
\end{description}

\subsection{{\tt{sor}}:
	Successive Over-Relaxation
	\label{s:toys-sor}}

This module solves a matrix equation $AX=V$ for
a dense, symmetric, diagonally dominant matrix $A$
and an arbitrary vector non-zero $V$
using successive over-relaxation.
Its inputs are:
\begin{description}
\item[{\tt{matrix}}:]
	the real matrix $A$.
\item[{\tt{target}}:]
	the real vector $V$.
\item[{\tt{nelts}}:]
	the number of values in each vector, and the size of the matrix along each axis.
\item[{\tt{tolerance}}:]
	the solution tolerance, e.g.\ $10^{-6}$.
\end{description}
Its output is:
\begin{description}
\item[{\tt{solution}}:]
	a real vector containing the solution $X$.
\end{description}

\subsection{{\tt{product}}:
	Matrix-Vector Product
	\label{s:toys-product}}

Given a matrix $A$,
a vector $V$,
and an assumed solution $X$ to the equation $AX=V$,
this module calculates the actual product $AX={V^{\prime}}$,
and then finds the magnitude of the error.
Inputs are:
\begin{description}
\item[{\tt{matrix}}:]
	the real matrix $A$.
\item[{\tt{actual}}:]
	the real vector $V$.
\item[{\tt{candidate}}:]
	a real vector containing the supposed solution.
\item[{\tt{nelts}}:]
	the number of values in each vector, and the size of the matrix along each axis.
\end{description}
The output of this function is:
\begin{description}
\item[{\tt{e}}:]
	the largest absolute value in the element-wise difference of $V$ and $V^{\prime}$.
\end{description}

\section{Other Issues\label{s:issues}}

As described in Section~\ref{s:method-softeng},
implementors are required to chain modules together in the manner shown in Figure~\ref{f:chaining}.
This section describes the flow of control and data between these chained problems,
and discusses several other important aspects of this suite,
including I/O and visualization.

\subsection{Chained Problem Sequences\label{s:issues-chain}}

The chained implementation of these problems
executes individual toys in the following order:
\begin{enumerate}
\item	An integer matrix $I$ with $r$ rows and $c$ columns is created
	either by using the Mandelbrot Set algorithm (Section~\ref{s:toys-mandel}) or by filling locations with random values (Section~\ref{s:toys-rng}).
	\label{e:chain-choice-1}
\item	The integer matrix $I$ is shuffled in both dimensions (Section~\ref{s:toys-half}).
\item	Either invasion percolation (Section~\ref{s:toys-invperc}) or histogram thresholding (Section~\ref{s:toys-thresh}) is used
	to generate a Boolean mask $B$ from $I$ in which the (minimum) percentage of {\tt{true}} locations is $P$.
	Like $I$, $B$ has $r$ rows and $c$ columns.
	\label{e:chain-choice-2}
\item	The Game of Life (Section~\ref{s:toys-life}) is simulated for $G$ generations, using $B$ as an initial configuration.
	This step overwrites the Boolean matrix $B$.
\item	A vector of $L$ $(m,x,y)$ points is created using the integer matrix $I$ and the Boolean matrix $B$ as described in Section~\ref{s:toys-winnow}.
\item	The elastic net algorithm is simulated using the points created in the previous step as cities.
\item	The final locations of the net points from the prevoius step are normalized, as described in Section~\ref{s:toys-norm}.
\item	An $L{\times}L$ matrix $A$ and an $L$ vector $V$ are created using the normalized point locations from the previous step,
	as described in Section~\ref{s:toys-outer}.
\item	The matrix equation $AX=V$ is solved using Gaussian elimination and using successive over-relaxation
	to generate two solution vectors $X_{gauss}$ and $X_{sor}$.
	In a parallel system, these two problems should if possible be executed concurrently.
\item	The checking vectors $V_{gauss} = A X_{gauss}$ and $V_{sor} = A X_{sor}$ are calculated.
	In a parallel system, these two problems should if possible be executed concurrently.
\item	The norm-1 distance between $V_{gauss}$ and $V_{sor}$, i.e.\ the greatest element-wise absolute difference, is calculated.
	This measures the agreement between the solutions found by the two methods.
\end{enumerate}
Because of choices in steps~\ref{e:chain-choice-1} and~\ref{e:chain-choice-2},
there are four possible chained sequences.
A minimal implementation will provide one of these,
while a full implementation will be a single program which can execute any of the four sequences.
Intermediate implementations providing all four sequences as separate programs are also acceptable.

\subsection{Input and Output\label{s:issues-io}}

>>>>

@@

\begin{figure}
% \epsfysize=0.5\textheight
% \begin{center}\mbox{\epsffile{fig/taxonomy.eps}}\end{center}
\caption{A Taxonomy of Parallel Programming Systems\label{f:taxonomy}}
\end{figure}

@@

\begin{figure}
% \epsfxsize=12cm
% \begin{center}\mbox{\epsffile{fig/chain.eps}}\end{center}
\caption{Chaining\label{f:chaining}}
\end{figure}

@@

\begin{figure}
% \epsfxsize=12cm
% \begin{center}\mbox{\epsffile{fig/perf-time.eps}}\end{center}
\caption{Performance vs.\ Time\label{f:perf}}
\end{figure}

@@

\begin{figure}
% \epsfxsize=6cm
% \begin{center}\mbox{\epsffile{fig/perf-complex.eps}}\end{center}
\caption{Comparative Code Complexity\label{f:complex}}
\end{figure}

@@
